---
title: "Boxplot Exercises"
author: "Stephen Blatti"
date: "August 8, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
The InsectSprays data set measures the counts of insects in agricultural experimental units treated with different insecticides. This dataset is included in R, and you can examine it by typing:

head(InsectSprays)

Try out two equivalent ways of drawing boxplots in R, using the InsectSprays dataset. Below is pseudocode, which you should modify to work with the InsectSprays dataset.

    1) using split:
    
    boxplot(split(values, factor))
    
    2) using a formula:

boxplot(values ~ factor)

1. Which spray seems the most effective (has the lowest median)?
```{r}
dat <- InsectSprays
head(dat)
boxplot(split(dat$count, dat$spray))
boxplot(dat$count ~ dat$spray)
```

2. Let's consider a random sample of finishers from the New York City Marathon in 2002. This dataset can be found in the UsingR package. Load the library and then load the nym.2002 dataset.

    library(dplyr)
    data(nym.2002, package="UsingR")

Use boxplots and histograms to compare the finishing times of males and females. Which of the following best describes the difference? 

```{r}
#install.packages(dplyr)
library(dplyr)
#install.packages("UsingR")
#library(UsingR)
data(nym.2002, package="UsingR")
x <- filter(nym.2002, gender=="Male") %>% select(time) %>% unlist 
y <- filter(nym.2002, gender=="Female") %>% select(time) %>% unlist 

par(mfrow = c(1,3))
boxplot(y, x)
hist(y,xlim=c(range(nym.2002$time)))
hist(x,xlim=c(range(nym.2002$time)))
```














