---
title: "QQ-plot Exercises"
author: "Stephen Blatti"
date: "August 8, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Download this RData file to your working directory: link. Then load the data into R with the following command:

load("skew.RData")


You should have a 1000 x 9 dimensional matrix 'dat':

dim(dat)

Using QQ-plots, compare the distribution of each column of the matrix to a normal. That is, use qqnorm() on each column. To accomplish this quickly, you can use the following line of code to set up a grid for 3x3=9 plots. ("mfrow" means we want a multifigure grid filled in row-by-row. Another choice is mfcol.)

par(mfrow = c(3,3))

Then you can use a for loop, to loop through the columns, and display one qqnorm() plot at a time. You should replace the text between ** with your own code.

for (i in 1:9) {
  **put your qqnorm call here**
}

Identify the two columns which are skewed.

Examine each of these two columns using a histogram. Note which column has "positive skew", in other words the histogram shows a long tail to the right (toward larger values). Note which column has "negative skew", that is, a long tail to the left (toward smaller values). Note that positive skew looks like an up-shaping curve in a qqnorm() plot, while negative skew looks like a down-shaping curve.

You can use the following line to reset your graph to just show one at a time:
par(mfrow=c(1,1))

```{r}
load("skew.RData")
dim(dat)
par(mfrow = c(3,3))
for (i in 1:9) {
  x <- dat[,i]
  qqnorm(x, main = paste0("Column=", i), ylim=c(-6,6))
  qqline(x)
}
par(mfrow=c(1,2))
hist(dat[,4])
hist(dat[,9])
```































