---
title: "Symmetry of Log Ratios Exercises"
author: "Stephen Blatti"
date: "August 12, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

n the previous video, we saw that multiplicative changes are symmetric around 0 when we are on the logarithmic scale. In other words, if we use the log scale, 1/2 times a number x, and 2 times a number x, are equally far away from x. We will explore this with the NYC marathon data.

Create a vector time of the sorted times:

```{r}
library(dplyr)
data(nym.2002, package="UsingR")
time = sort(nym.2002$time)
```

1. What is the fastest time divided the median time?
```{r}
head(time)
time[1] / median(time)
# or min(time) / median(time)
```
2. What is the slowest time divided the median time?
```{r}
max(time) / median(time)
```

              
        
        Compare the following two plots.
        
        1) A plot of the ratio of times to the median time, with horizontal lines at twice as fast as the median time, and twice as slow as the median time.
        
        plot(time/median(time), ylim=c(1/4,4))
        abline(h=c(1/2,1,2))
        
        2) A plot of the log2 ratio of times to the median time. The horizontal lines indicate the same as above: twice as fast and twice as slow.
        
        plot(log2(time/median(time)),ylim=c(-2,2))
        abline(h=-1:1)
        
        Note that the lines are equally spaced in Figure #2.
        
        
```{r}
par(mfrow=c(1,2))
plot(time/median(time), ylim=c(1/4,4))
abline(h=c(1/2,1,2))
plot(log2(time/median(time)),ylim=c(-2,2))
abline(h=-1:1)
```












