---
title: "Error Rates and Procedures Exercises"
author: "Stephen Blatti"
date: "August 29, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
n this assessment we hope to help you further grasp the concept that p-values are random variables and start laying the ground work for the development of procedures that control error rates. The calculations to compute error rates require us to understand the random behavior of p-values.

We are going to ask you to perform some calculations related to introductory probability theory. One particular concept you need to grasp is statistical independence. You also will need to know that the probability of two random events that are statistically independent occurring is P( A and B) = P(A)P(B). Note that this is a consequence of the more general formula P(A and B) = P(A) P(B | A )

1. Assume the null is true and denote the p-value you would get if you ran a test as . Define the function What does

look like?
a uniform distribution
the identity line correct
a constant at 0.05
P is not a random value

2. In the previous assessment we saw how the probability of incorrectly rejecting the null for at least one of 20 experiments for which the null is true is well over 5%. Now let's consider a case in which we run thousands of tests as we would do in a high throughput experiment.

We previously learned that under the null, the probability of a p-value < p is p. If we run 8,793 independent tests, what is the probability of incorrectly rejecting at least one of the null hypotheses? 

```{r}
# N <- 1000
# minpval <- replicate(N, min(runif(8793,0,1))<0.05)
# mean(minpval >= 1)

alpha=0.05
1 - (1 - alpha)^8793
```

3. Suppose we need to run 8,793 statistical tests and we want to make the probability of a mistake very small, say 5%. Using the answer to exercise #2, how small do we have to change the cutoff, previously 0.05, to lower our probability of at least one mistake to be 5%. 

```{r}
# Trial and error
N <- 1000
minpval <- replicate(N, min(runif(8793,0,1))<0.0000058)
mean(minpval >= 1)
# answer 0.000005833407
# or this
1 - (1 - alpha)^(1/8793)

##warning this can take several minutes
##and will only give an approximate answer
# B=10000
# cutoffs = 10^seq(-7,-4,0.1) ##we know it has to be small
# prob = sapply(cutoffs,function(cutoff){
#     minpval =replicate(B, min(runif(8793,0,1))<=cutoff)
#     mean(minpval>=1)
#     })
# cutoffs[which.min(abs(prob-0.05))]
```






