---
title: "Association Tests Exercises"
author: "Stephen Blatti"
date: "August 7, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



In the previous video, Rafa showed how to calculate a Chi-square test from a table. Here we will show how to generate the table from data which is in the form of a dataframe, so that you can then perform an association test to see if two columns have an enrichment (or depletion) of shared occurances.

Download the assoctest.csv file into your R working directory, and then read it into R:
```{r}
d = read.csv("assoctest.csv")
```

This dataframe reflects the allele status (either AA/Aa or aa) and the case/control status for 72 individuals.

1. Compute the Chi-square test for the association of genotype with case/control status (using the table() function and the chisq.test() function). Examine the table to see if it look enriched for association by eye. What is the X-squared statistic?

```{r}
head(d)
tab <- table(d)
tab
chisq.test(tab)
# or sln 
#tab = table(d$allele, d$case)
#chisq.test(tab)

```

2. Compute the Fisher's exact test ( fisher.test() ) for the same table. What is the p-value?
```{r}
fisher.test(tab)
```







