---
title: "Populations, Samples, & Estimates"
author: "Stephen Blatti"
date: "July 31, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Populations, Samples, and Estimates
```{r}
library(downloader) 
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/mice_pheno.csv"
filename <- basename(url)
download(url, destfile=filename)
dat <- read.csv(filename) 
dat <- na.omit(dat)
```
1. Use dplyr to create a vector x with the body weight of all males on the control (chow) diet. 
What is this population's average? 
```{r}
library(dplyr)
head(dat)
x <- filter(dat, Sex %in% "M" & Diet == "chow") %>%
  select(Bodyweight) %>% 
  unlist
head(x)
mean(x)
```
2. Now use the rafalib package and use the popsd function to compute the population standard deviation. 
```{r}
# install.packages("rafalib")
library(rafalib)
popsd(x, na.rm = TRUE)
```
3. Set the seed at 1. 
Take a random sample X of size 25 from x. What is the sample average? 
```{r}
set.seed(1)
rs_X <- sample(x, 25)
mean(rs_X)
```
4. Use dplyr to create a vector y with the body weight of all males on the high fat hf) diet. 
What is this population's average? 
```{r}
y <- filter(dat, Sex %in% "M" & Diet == "hf") %>%
  select(Bodyweight) %>% 
  unlist
mean(y)
```
5. Now use the rafalib package and use the popsd function to compute the population standard deviation.
```{r}
popsd(y)
```
6. Set the seed at 1. Take a random sample of size 25 from y. What is the sample average?
```{r}
set.seed(1)
rs_Y <- sample(y, 25)
mean(rs_Y)
```
7. What is the difference in absolute value between mean(y) - mean(x) and mean(rs_y) - mean(rs_X)? 
```{r}
abs((mean(y) - mean(x)) - (mean(rs_Y) - mean(rs_X)))
```
8. Repeat the above for females. Make sure to set the seed to 1 before each sample call. What is the difference in absolute value between mean(y) - mean(x) and mean(rs_y) - mean(rs_X)? 
```{r}
x <- filter(dat, Sex %in% "F" & Diet == "chow") %>%
  select(Bodyweight) %>% 
  unlist
y <- filter(dat, Sex %in% "F" & Diet == "hf") %>%
  select(Bodyweight) %>% 
  unlist
set.seed(1)
rs_X <- sample(x, 25)
set.seed(1)
rs_Y <- sample(y, 25)
abs((mean(y) - mean(x)) - (mean(rs_Y) - mean(rs_X)))
```
9. For the females, our sample estimates were closer to the population difference than with males. What is a possible explanation for this?
a. The population variance of the females is smaller than that of the males; thus, the sample variable has less variability. correct
b. Statistical estimates are more precise for females.
c. The sample size was larger for females.
d. The sample size was smaller for females.
```{r}
popsd(x)
popsd(y)
```


