---
title: "Models for Variance Exercises"
author: "Stephen Blatti"
date: "September 11, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Install and load the following data library:
```{r}
library(devtools)
install_github("genomicsclass/tissuesGeneExpression")
library(tissuesGeneExpression)

# Now load this data and select the columns related to endometrium: 

data("tissuesGeneExpression")
library(genefilter)
y = e[,which(tissue=="endometrium")]
```
This will give you a matrix y with 15 samples.

1. Compute the across sample variance for the fifteen samples. Then make a qq-plot to see if the variances follow ae normal distribution.
```{r}
s2 <- rowVars(y) # between sample variances
datsds<-rowSds(y) # between sample standard deviations
library(rafalib)
mypar(1,2)
qqnorm(s2)
qqline(s2)
# To see the square root transformation does not help much:
qqnorm(sqrt(s2))
qqline(sqrt(s2))

```
Which statement is true? (pick one)

    With the exception of a handful of outliers, the data follow a normal distribution.
    The variance does not follow a normal distribution, but taking the square root fixes this.
    The normal distribution is not a useful approximation here: the left tail is over estimated and the right tail is underestimated. correct
    The normal distribution fits the data almost perfectly. 

2. Now fit an F-distribution with 14 degrees of freedom using the fitFDist function in the limma package:

What is estimated the estimated scale parameter?
```{r}
# p 295
library(limma)
estimates <- fitFDist(datsds^2,14)
estimates$scale
# or estimates=fitFDist(s2,14)

```
3. Now create a qq-plot of the observed sample standard deviation versus the quantiles predicted by the F-distribution (remember to take square root).

```{r}
#qq plots
ps <- (seq(along=s2)-0.5)/length(s2)
theoretical<- qf(ps,14,estimates$df2)*estimates$scale 
LIM <- sqrt( range(c(theoretical,s2)) )
mypar(1,2)
qqplot(sqrt( theoretical ), sqrt( s2 ),ylim=LIM,xlim=LIM)
abline(0,1)

##close up excluding the upper 5% gives good fit
K <- sqrt( quantile(s2,0.95) )
qqplot( sqrt( theoretical ), sqrt( s2 ),ylim=c(0,K),xlim=c(0,K))
abline(0,1)

#histogram p 296
sds=seq(0,2,len=100)
tmp=hist(observed,main=paste("s_0 =", signif(estimates[[1]],2), "d =",
         signif(estimates[[2]],2)), xlab="sd", ylab="density",
         freq=FALSE, nc=100, xlim=c(0,2), ylim=c(0,4))
dd=df(sds^2/estimates$scale,14,estimates$df2)
k=sum(tmp$density)/sum(dd) ##a normalizing constant to assure same area in plot
lines(sds, dd*k, type="l", col=2, lwd=2)


```

Which of the following best describes the qq-plot?

    The fitted F-distribution provides a perfect fit.
    If we exclude the genes with the highest variances (top 5%), the F-distribution provides a good fit. correct
    If we exclude the genes with the lowest variances (bottom 5%), the F-distribution provides a good fit.
    The normal distribution provided a better fit. 



























