---
title: "Projections Exercises"
date: "September 20, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
We will continue to use this dataset:
```{r}
library(Biobase)
library(GSE5859Subset)
data(GSE5859Subset)
```

1.Suppose you want to make an MA plot of the first two samples y = geneExpression[,1:2]. Which of the following projections of gives us new coordinates such that column 2 versus column 1 is an MA plot?

$$ y\left(\begin{array}{ccc}
1/\sqrt(2) & 1/\sqrt(2) & 1/\sqrt(2)	\\
1/\sqrt(2) & 1/\sqrt(2) & 1/\sqrt(2)	\\
1/\sqrt(2) & 1/\sqrt(2) & -1/\sqrt(2)	\\
\end{array}\right)
$$

$$ y\left(\begin{array}{ccc}
1 & 1 & 1	\\
1 & 1 & 1	\\
1 & 1 & -1	\\
\end{array}\right)
$$

$$ \left(\begin{array}{ccc}
1 & 1 & 1	\\
1 & 1 & 1	\\
1 & 1 & -1	\\
\end{array}\right)y
$$

$$ \left(\begin{array}{ccc}
1 & 1 & 1	\\
1 & 1 & 1	\\
1 & 1 & -1	\\
\end{array}\right)y^t
$$
The 2nd one is correct.

2. Say Y is M x N, in the SVD $Y = UDV^T $ which of the following is NOT correct?

$DV^T$ are the new coordinates for the projection $U^TY$ <br />
UD are the new coordinates for the projection YV <br />
D are the coordinates of the projection $U^TY$ correct <br />
$U^TY$ is a projection from an M- dimensional to N - dimensional subspace <br />





