---
title: "SVD Exercises"
date: "September 20, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


For the following questions use the data loaded with:
```{r}
library(tissuesGeneExpression)
data(tissuesGeneExpression)
```
Important note: When using the SVD in practice it is important to note that the solution to SVD is not unique. This is because $UDV^T = (-U)D(-V)^T$. In fact we can flip the sign of each column of U and as long as we also flip the respective column in V the decompostion works. Here R code demonstrating this:
```{r}
s = svd(e)
signflips = sample(c(-1,1),ncol(e),replace=TRUE)
signflips
```
Now we switch the sign of each column and check that we get the same answer. We do this using the function sweep. If x is a matrix and a is a vector, then sweep(x,1,y,FUN="*") applies the fun FUN to each row i FUN(x[i,],a[i]), in this case x[i,]*a[i]. If instead of 1 we use 2 sweep applies this to columns. To learn about sweep read ?sweep.
```{r}
newu= sweep(s$u,2,signflips,FUN="*")
newv= sweep(s$v,2,signflips,FUN="*" )
all.equal( s$u %*% diag(s$d) %*% t(s$v), newu %*% diag(s$d) %*% t(newv))
```
This is important to know because different implementations of the SVD algorithm may give different signs, which can lead to the same code resulting in different answers when run in different computer systems.

1.  Compute the SVD of e
```{r}
s = svd(e)

# Now compute the mean of each row:

m = rowMeans(e)
```
What is the correlation between the first column of
and m?
```{r}
cor(m,s$u[,1])
```
2. In the above question we saw how the first column relates to the mean of the rows of e. Note that if we change these means, the distances between columns do not change. Here's some R code showing how changing the means does not change the distances:
```{r}
newmeans = rnorm(nrow(e)) ##random values we will add to create new means
newe = e+newmeans ##we change the means
sqrt(crossprod(e[,3]-e[,45]))
sqrt(crossprod(newe[,3]-newe[,45])) 
``` 

So we might as well make the mean of each row 0 since it does not help us approximate the column distances. We will define y as the detrended e and recompute the SVD:
```{r}
y = e - rowMeans(e)
s = svd(y)
```
We showed that $UDV^T$ is equal to y up to numerical error
```{r}
resid = y - s$u %*% diag(s$d) %*% t(s$v)
max(abs(resid))
```
The above can be made more efficient in two ways. First, using the crossprod and second not creating a diagonal matrix. Note that in R we can multiply a matrix x by vector a. The result is a matrix with row i equal to x[i,]*a[i]. Here is an example to illustrate this.
```{r}
x=matrix(rep(c(1,2),each=5),5,2)
x

x*c(1:5)
```
Note that the above code is actually equivalent to:
```{r}
sweep(x,1,1:5,"*")
```

This means that we don't have to convert s\$d into a matrix to obtain $DV^T$ 

Which of the following gives us the same as diag(s$d)%*%t(s$v)

    s$d %*% t(s$v)1
    s$d * t(s$v)
    t(s$d * s$v)
    s$v * s$d

```{r}
#diag(s$d)%*%t(s$v)
head(s$d * t(s$v)[,1:5] )

```

3. If we define vd = t(s$d * t(s$v)) then which of the following is not the same as $UDV^T$

    tcrossprod(s$u,vd)
    s$u %*% s$d * t(s$v)
    s$u %*% (s$d * t(s$v) )
    tcrossprod( t( s$d*t(s$u)) , s$v)


```{r}
vd = t(s$d * t(s$v))
#head(s$u %*% s$d * t(s$v)[,1:5])
#Order of operations in R go left to right and s$u %*% s$d is multiplying non-conformable matrices

```
4. Let z = s$d * t(s$v). We showed derivation demonstrating that because U is orthogonal the distance between e[,3] and e[,45] is the same as the distance between y[,3] and y[,45] which is the same as z[,3] and z[,45]
```{r}
z = s$d * t(s$v)

sqrt(crossprod(e[,3]-e[,45]))
sqrt(crossprod(y[,3]-y[,45]))
sqrt(crossprod(z[,3]-z[,45]))
```
Note that the columns z have 189 entries, compared to 22,215 for e.

What is the difference (in absolute value) between the actual distance sqrt(crossprod(e[,3]-e[,45])) and the approximation using only two dimension of z

```{r}
dim(z)

actual_dist<-sqrt(crossprod(e[,3]-e[,45]))
approx_dist<-sqrt(crossprod(z[1:2,3]-z[1:2,45]))

abs(actual_dist - approx_dist)
```
5. What is the minium number of dimensions we need to use for the approximation in SVD Exercises #4 to be within 10% or less?
```{r}
ks = 1:189
actual_distance = sqrt(crossprod(e[,3] - e[,45]))
approx_distances = sapply(ks,function(k){
    sqrt(crossprod(z[1:k,3,drop=FALSE]-z[1:k,45,drop=FALSE] )) 
})
percent_diff = 100 * abs(approx_distances - actual_distance) / actual_distance
plot(ks,percent_diff) 
min(ks[which(percent_diff < 10)])
```
6. Compute distances between sample 3 and all other samples:

    distances = sqrt(apply(e[,-3]-e[,3],2,crossprod))

Recompute this distance using the 2 dimensional approximation.
What is the Spearman correlation between this approximate distance and the actual distance?
```{r}
distances = sqrt(apply(e[,-3]-e[,3],2,crossprod))
approx_distances = sqrt(apply(z[1:2,-3]-z[1:2,3],2,crossprod))
plot(distances,approx_distances) 
cor(distances,approx_distances,method = "spearman")
```



















