---
title: "Exploratory Data Analysis Exercises"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Download and install the Bioconductor package SpikeInSubset and then load the library and the mas133 data
```{r}
source("http://www.bioconductor.org/biocLite.R")
biocLite("SpikeInSubset")
library(SpikeInSubset)
data(mas133)
 

# Now make the following plot of the first two samples and compute the correlation:

e <- exprs(mas133)
plot(e[,1],e[,2],main=paste0("corr=",signif(cor(e[,1],e[,2]),3)),cex=0.5)
k <- 3000
b <- 1000 #a buffer
polygon(c(-b,k,k,-b),c(-b,-b,k,k),col="red",density=0,border="red")
```

1. What proportion of the points are inside the box?
```{r}
# Want to know which genes are below k for both samples
mean(e[,1]<k & e[,2]<k)
```
2. Now make the sample plot with log:
```{r}
plot(log2(e[,1]),log2(e[,2]),main=paste0("corr=",signif(cor(log2(e[,1]),log2(e[,2])),2)),cex=0.5)
k <- log2(3000)
b <- log2(0.5)
polygon(c(b,k,k,b),c(b,b,k,k),col="red",density=0,border="red")
# When you take the log, 95% of data is no longer in a tiny section of plot.
```

What is an advantage of taking the log?

    The tails do not dominate the plot. corrext
    There are less points.
    There is exponential growth.
    We always take logs.


3. Make an MA-plot
```{r}
e <- log2(exprs(mas133))
plot((e[,1]+e[,2])/2,e[,2]-e[,1],cex=0.5)
```
The two samples we are plotting are replicates (they random samples from the same batch of RNA) The correlation of the data was 0.997 in the original scale, 0.96 in the log-scale. High correlations are sometimes confused for evidence of replication. But replication implies we get very small difference between the observations which is better measured with distance or differences.
What is the standard deviation of the log ratios for this comparison?
```{r}
log_ratio <- (e[,1]-e[,2])
sd(log_ratio)


# However, note that if there is a mean shift the standard deviation will not summarize this. We can instead consider the average distance: 
sqrt(mean( (e[,2]-e[,1])^2) )

```

4. How many fold changes above 2 do we see? Note that these measures of log (base 2) of expression so a fold change of 2 translates into a difference, in absolute value, of 1. 
```{r}
sum(abs(log_ratio) > 1)

#  This are log2 measurements so a fold-change of 2 relates to differences of 1 (in absolute value). We then simply count the occurrences
```














