---
title: "Contrasts Exercises"
author: "Stephen Blatti"
date: "August 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Suppose we have an experiment with two species A and B, and two conditions: control and treated.

species <- factor(c("A","A","B","B"))
condition <- factor(c("control","treated","control","treated"))

And we will use a formula of '~ species + condition'.

The model matrix is then:

model.matrix(~ species + condition)

1. Suppose we want to build a contrast of coefficients for the above experimental design.

You can either figure this question out through logic, by looking at the design matrix, or using the contrast() function from the contrast library. If you have not done so already, you should download the contrast library. The contrast vector is returned as contrast(...)$X.

What should the contrast vector be, for the contrast of (species=B and condition=control) vs (species=A and condition=treatment)? Assume that the beta vector from the model fit by R is: Intercept, speciesB, conditiontreated.

```{r}
# not 011
library(contrast)
species <- factor(c("A","A","B","B"))
condition <- factor(c("control","treated","control","treated"))
X <- model.matrix(~ species + condition)
X
# So we have 1 1 0
#         -  1 0 1
# gives      0 1 -1
```
As you want to compare species B vs A and condition control vs treated, you want +1 for the speciesB coefficient and -1 for the conditiontreated coefficient.

Another way to find this would be by generating some random y, fitting a model, and using contrast():

y = rnorm(4)

fit = lm(y ~ species + condition)

contrast(fit, list(species="B",condition="control"), list(species="A",condition="treated"))$X


2. Load the spider dataset like this:
```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/spider_wolff_gorb_2013.csv"
filename <- "spider_wolff_gorb_2013.csv"
library(downloader)
if (!file.exists(filename)) download(url, filename)
spider <- read.csv(filename, skip=1)
```
Use the Rmd script of the spider dataset. Suppose we build a model using two variables: ~ type + leg.
What is the t-value for the contrast of leg pair L4 vs leg pair L2?
```{r}
fit1 <- lm(friction ~ type + leg, data = spider)
(L4vsL2 <- contrast(fit1, llist(leg="L4",type="pull"),list(leg="L2",type="pull")) )

```

3. The t-value for the contrast of leg pair L4 vs leg pair L2 is constructed by taking the difference of the coefficients legL4 and legL2, and then dividing by the standard error of the difference. In the last question we will explore how the standard error of the difference is calculated here.

In the book page for contrasts (see the heading, Contrasting the coefficients), we saw that in general, for a contrast vector C, the standard error of the contrast (CB_hat) is: sqrt( C sum(C^T))

Sigma, sumation, is the covariance matrix of beta-hat, B_hat. The covariance matrix contains elements which give the variance or covariance of elements in beta-hat. The elements on the diagonal of the Sigma matrix give the variance of each element in beta-hat. The square root of these is the standard error of the elements in beta-hat. The off-diagonal elements of Sigma give the covariance of two different elements of the beta-hat matrix. So Sigma[1,2] gives the covariance of the first and second element of beta-hat. The Sigma matrix is symmetric, which means Sigma[i,j] = Sigma[j,i].

For the difference, we have that:
var( beta^_L4-  beta^_L2) = var( beta^_L4) + var( beta^_L2) - 2cov(beta^_L4, beta^_L2)

But we can also work out in this simple case, where we simply subtract one coefficient from another, using the formula for the variance of sums of random variables:

In the book page, we computed Sigma using:

X <- model.matrix(~ type + leg, data=spider)
(Sigma <- sum(fitTL$residuals^2)/(nrow(X) - ncol(X)) * solve(t(X) %*% X))

Our contrast matrix is:

C <- matrix(c(0,0,-1,0,1),1,5)


Using Sigma, what is Cov(beta-hat_L4, beta-hat_L2)?
```{r}
X <- model.matrix(~ type + leg, data=spider)
(Sigma <- sum(fit1$residuals^2)/(nrow(X) - ncol(X)) * solve(t(X) %*% X))
C <- matrix(c(0,0,-1,0,1),1,5)
Sigma[3,5]
```



Confirm that 

sqrt(Var(beta-hat_L4 - beta-hat_L2)) = sqrt(Var(beta-hat_L4) + Var(beta-hat_L2) - 2 Cov(beta-hat_L4, beta-hat_L2))

is equal to

sqrt(C %*% Sigma %*% t(C))

is equal to the standard error from the contrast() for the leg L4 vs L2 difference.







