---
title: "Matrix Algebra in Practice Exercises"
author: "Stephen Blatti"
date: "August 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Suppose we are analyzing a set of 4 samples. The first two samples are from a treatment group A and the second two samples are from a treatment group B. This design can be represented with a model matrix like so:

X <- matrix(c(1,1,1,1,0,0,1,1),nrow=4)
rownames(X) <- c("a","a","b","b")

X will then look like:

a 1 0
a 1 0
b 1 1
b 1 1

Suppose that the fitted parameters for a linear model give us:

beta <- c(5, 2)

Use the matrix multiplication operator, %*%, in R to answer the following questions:

1. What is the fitted value for the A samples? (The fitted Y values.)

```{r}
X <- matrix(c(1,1,1,1,0,0,1,1),nrow=4)
rownames(X) <- c("a","a","b","b")
beta <- c(5, 2)
fitted = X %*% beta
fitted[1:2, ]
```

2. What is the fitted value for the B samples? (The fitted Y values.)

```{r}
fitted[3:4, ]
```

Suppose now we are comparing two treatments B and C to a control group A, each with two samples. This design can be represented with a model matrix like so:

X <- matrix(c(1,1,1,1,1,1,0,0,1,1,0,0,0,0,0,0,1,1),nrow=6)
rownames(X) <- c("a","a","b","b","c","c")

which results in a matrix that looks like

a 1 0 0
a 1 0 0
b 1 1 0
b 1 1 0
c 1 0 1
c 1 0 1

Suppose that the fitted values for the linear model are given by:

beta <- c(10,3,-3)

3. What is the fitted value for the B samples?

```{r}
X <- matrix(c(1,1,1,1,1,1,0,0,1,1,0,0,0,0,0,0,1,1),nrow=6)
rownames(X) <- c("a","a","b","b","c","c")
beta <- c(10,3,-3)
fitted = X %*% beta
fitted[3:4, ]
```

4. What is the fitted value for the C samples?

```{r}
fitted[5:6, ]
```






















