---
title: "Matrix Operation Exercises"
author: "Stephen Blatti"
date: "August 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Suppose X is a matrix in R. Which of the following is NOT equivalent to X?

          t( t(X) )
          X %*% matrix(1,ncol(X) ) correct
          X*1
          X%*%diag(ncol(X)) 

```{r}
r1 <- c(1,3,2)
r2 <- c(1,-2,1)
r3 <- c(1,1,-1)
abc <- c(6,2,1)
X <- matrix(cbind(r1,r2,r3),3,3)
Y <- matrix(abc,3,1)
solve(X,Y)
X
t( t(X) )
X %*% matrix(1,ncol(X) )
X*1
X%*%diag(ncol(X)) 
```
Explanation

The first transposes the transpose, so we end up with our original X. The third is multiplying each element by 1, and the fourth is multiplying X by the identity. The second is not even guaranteed to have the same dimensions as X.

2. Solve the following system of equations using R:

        3a + 4b - 5c + d = 10
        
        2a + 2b + 2c - d = 5
        
        a -b + 5c - 5d = 7
        
        5a + d = 4

What is the solution for c:

```{r}
r1 <- c(3,4,-5,1)
r2 <- c(2,2,2,-1)
r3 <- c(1,-1,5,-5)
r4 <- c(5,0,0,1)
abcd <- c(10,5,7,4)
X <- matrix(rbind(r1,r2,r3,r4),4,4)
X
solve(X, abcd)
```

3. Load the following two matrices into R:

a <- matrix(1:12, nrow=4)
b <- matrix(1:15, nrow=3)

Note the dimension of 'a' and the dimension of 'b'.

In the question below, we will use the matrix multiplication operator in R, %*%, to multiply these two matrices.

What is the value in the 3rd row and the 2nd column of the matrix product of 'a' and 'b'?

```{r}
a <- matrix(1:12, nrow=4)
b <- matrix(1:15, nrow=3)
c <- a %*% b
c[3,2]

```
4. Multiply the 3rd row of 'a' with the 2nd column of 'b', using the element-wise vector multiplication with *.
What is the sum of the elements in the resulting vector?

```{r}
sum(a[3,] * b[,2])
```






















