---
title: "Matrix Notation Exercises"
author: "Stephen Blatti"
date: "August 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. In R we have vectors and matrices. You can create your own vectors with the function c.

c(1,5,3,4)

They are also the output of many functions such as

rnorm(10)

You can turn vectors into matrices using functions such as rbind, cbind or matrix.

Create the matrix from the vector 1:1000 like this:

X = matrix(1:1000,100,10)
What is the entry in row 25, column 3 ?

```{r}
X = matrix(1:1000,100,10)
X[25,3]
```


2. Using the function cbind, create a 10 x 5 matrix with first column x=1:10. Then columns 2*x, 3*x, 4*x and 5*x in columns 2 through 5.
What is the sum of the elements of the 7th row?

```{r}
x<- 1:10
X <- cbind(x, 2*x, 3*x, 4*x, 5*x)
sum(X[7,])
```

3. Which of the following creates a matrix with multiples of 3 in the third column?

        matrix(1:60,20,3)
        matrix(1:60,20,3,byrow=TRUE) correct
        x=11:20;rbind(x,2*x,3*x)
        x=1:40;matrix(3*x,20,2)

```{r}
matrix(1:60,20,3,byrow=TRUE) 
```
Explanation

You can make each of the matrices in R and examine them visually. Or you can check whether the third column has all multiples of 3 with all(X[,3]%%3==0). Note that the fourth choice does not even have a 3rd column.


















