---
title: "Hiearchichal Clustering Exercises"
date: "September 25, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
1. Create a random matrix with no correlation in the following way:
```{r}
set.seed(1)
m = 10000
n = 24
x = matrix(rnorm(m*n),m,n)
colnames(x)=1:n
```

Run hierarchical clustering on this data with the hclust function with default parameters to cluster the columns. Create a dendrogram.
From the dendrogram which pairs of samples are the furthest away from each other?

    7 and 23
    19 and 14 correct
    1 and 16
    17 and 18 

```{r}
d <- dist( t(x) )
hc <- hclust(d)
plot(hc)
```

2. Set the seed at 1, set.seed(1) and replicate the creation of this matrix 100 times
    
    m = 10000
    n = 24
    x = matrix(rnorm(m*n),m,n)

then perform hierarchical clustering as in the solution to question 2.4.1 and find the number of clusters if you use cutree at height 143. Note that this number is a random variable.
Based on the Monte Carlo simulation, what is the standard error of this random variable?


```{r}
library(rafalib)
set.seed(1)
m = 10000
n = 24
N = 100
sim2 = replicate(N,{
  x = matrix(rnorm(m*n),m,n)
  d <- dist( t(x) )
  hc <- hclust(d)
  max(cutree(hc,h=143))
  
})
hist(sim2)
sd(sim2)



# Model Solution
# set.seed(1)
# m = 10000
# n = 24
# nc = replicate(100,{
#     x = matrix(rnorm(m*n),m,n)
#     hc = hclust( dist( t(x)))
#     length(unique(cutree(hc,h=143)))
# })
# plot(table(nc)) ## look at the distribution
# popsd(nc) # unbiased estimate of sd
```



















