---
title: "K-means Clustering Exercises"
date: "September 30, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Run kmeans with 4 centers for the blood RNA data:
```{r}
library(GSE5859Subset)
data(GSE5859Subset)
```

Set the seed to 10, set.seed(10) right before running kmeans with 5 centers.

Explore the relationship of clusters and information in sampleInfo. Which of the following best described what you find:

    sampleInfo$group is driving the clusters as the 0s and 1s are in completely different clusters/
    The year is driving the clusters
    Date is driving the clusters correct
    The clusters don't depend on any of the column of sampleInfo 

```{r}
set.seed(10)
d = dist(t(geneExpression))
km = kmeans(t(geneExpression), centers = 5)
mds = cmdscale(d)
plot(mds,bg=km$cluster,pch=21)
table(sampleInfo$group,km$cluster)
table(sampleInfo$date,km$cluster)
#looks better if we re-order:
table(sampleInfo$date,km$cluster)[,c(4,1,5,3,2)]

```










