---
title: "Smoothing Exercises"
date: "October 2, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
1. Use the data generated in question 2.6.2
```{r}
n = 10000
set.seed(1)
men = rnorm(n,176,7) #height in centimeters
women = rnorm(n,162,7) #height in centimeters
y = c(rep(0,n),rep(1,n))
x = round(c(men,women))
##mix it up
ind = sample(seq(along=y))
y = y[ind]
x = x[ind]
  

#Set the seed at 5, set.seed(5) and take a random sample of 250 individuals from the population like this:

set.seed(5)
N = 250
ind = sample(length(y),N)
Y = y[ind]
X = x[ind]
```
Use loess to estimate $f(x) = E(Y|X=x)$ using the default parameters. What is the predicted f(168)?
```{r}
fit <- loess(Y~X)
newx <- seq(160,178)
smooth <- predict(fit,newdata=data.frame(X=newx))
P_ys =sapply(newx,function(k){
  mean(Y[X==k])
  })

plot(newx,P_ys)
lines(newx,smooth)

smooth[newx==168]
predict(fit,newdata=data.frame(X=168))
```

2. The loess estimate above is a random variable thus we should compute its standard error. Use Monte Carlo simulation to compute the standard error of your estimate of f(168).

Set the seed to 5, set.seed(5) and perform 1000 simulation of the computations performed in question 2.7.1. Report the the SE of the loess based estimate.
```{r}
N_sim = 1000
xs = seq(160,178)
set.seed(5)
sim2 <- replicate(N_sim,{
  N <- 250
  ind <- sample(length(y),N)
  Y <- y[ind]
  X <- x[ind]
  fit <-loess(Y~X)
  smooth_estimate <- predict(fit,newdata=data.frame(X=168))
  return(smooth_estimate)
  })
library(rafalib)
popsd(sim2)

# Model Answer
# ##plot plots are optional
# set.seed(5)
# B = 1000
# N = 250
# xs = seq(160,178)
# plot(xs,xs,ylim=c(0,1),type="l")
# res = replicate(B,{
#   ind = sample(length(y),N)
#   Y = y[ind]
#   X = x[ind]
#   fit=loess(Y~X)
#   ##optional plots
#   fitted=predict(fit,newdata=data.frame(X=xs))
#   lines(xs,fitted)
#   estimate = predict(fit,newdata=data.frame(X=168))
#   return(estimate)
#   })
# library(rafalib)
# popsd(res)
```


